Extending Height Above Nearest Drainage to Model Multiple Fluvial Sources in Flood Inundation Mapping Applications for the U.S. National Water Model

### 2022-09-15 manuscripts/owp_fim4_2022/sections/conclusions.tex
* latest updates * possible resubmission candidate

The main contribution and conclusion of this work centers around a fundamental limitation in HAND based FIM which is a failure to account for multiple possible sources of fluvial inundation since HAND only considers inundation from the nearest drainage line.
We illustrate that reducing the Horton-Strahler stream order of a HAND processing unit down to one enhances skill by significantly reducing false negatives at junctions of major streams.
In order to reduce stream order of the NWM stream network for HAND computation, we dissected the NWM network into two simpler units of singular Horton-Strahler stream order and mosaiced the resulting FIMs derived from each.
The NWM Mainstems (MS) stream network, which covers roughly 4-5\% of the NWM Full Resolution (FR) network, spans all established forecasting points in the Advanced Hydrologic Prediction System (AHPS) and downstream reaches.
The inundation from MS derived HAND is mosaiced together with the inundation of FR derived HAND.
Extending order reduction to the entire network, the Generalized Mainstem (GMS) technique discretizes the NWM FR network into level paths (LP) of unit stream order for HAND computation.
All LP based FIM derived from LP based HAND datasets are mosaiced together to form one seamless FIM.
Dissecting the stream network into regions of LPs with unit stream order is necessary because HAND has a ``nearest drainage'' limitation meaning it only accounts for riverine inundation sourced from the nearest drainage line.
In our evaluation of this technique, we observe that HAND based FIM improves in skill as we extend from nearest drainage inundation in FR to multiple drainage support in MS for only 4-5\% of the FR network.
Extending multiple drainage support to the entire FR network with GMS based HAND improves upon skill at around the same magnitude that MS improved upon FR.
Thus we conclude that deriving HAND with independent stream networks of unit Horton-Strahler stream order enhances the skill of FIM but offers diminishing returns as we extend from 4-5\% of the network with MS to 100\% of the network with GMS since deriving HAND and FIMs at these localized scale does add computational expense.

>>>>

The main contribution and conclusion of this work centers around a fundamental limitation in HAND based FIM which is a failure to account for multiple possible sources of fluvial inundation since HAND only considers inundation from the nearest drainage line.
We illustrate that reducing the Horton-Strahler stream order of a HAND processing unit down to one enhances skill by significantly reducing false negatives at junctions of major streams.
In order to reduce stream order of the NWM stream network for HAND computation, we dissected the NWM network into two simpler units of singular Horton-Strahler stream order and mosaiced the resulting FIMs derived from each.
The NWM Mainstems (MS) stream network, which covers roughly 4-5\% of the NWM Full Resolution (FR) network, spans all established forecasting points in the Advanced Hydrologic Prediction System (AHPS) and downstream reaches.
The inundation from MS derived HAND is mosaiced together with the inundation of FR derived HAND.
Extending order reduction to the entire network, the Generalized Mainstem (GMS) technique discretizes the NWM FR network into level paths (LP) of unit stream order for HAND computation.
All LP based FIM derived from LP based HAND datasets are mosaiced together to form one seamless FIM.
Dissecting the stream network into regions of LPs with unit stream order is necessary because HAND has a ``nearest drainage'' limitation meaning it only accounts for riverine inundation sourced from the nearest drainage line.
In our evaluation of this technique, we observe that HAND based FIM improves in skill as we extend from nearest drainage inundation in FR to multiple drainage support in MS for only 4-5\% of the FR network.
Extending multiple drainage support to the entire FR network with GMS based HAND improves skill at around the same magnitude that MS improved upon FR.
Thus we conclude that deriving HAND with independent stream networks of unit Horton-Strahler stream order enhances the skill of FIM but offers diminishing returns as we extend from 4-5\% of the network with MS to 100\% of the network with GMS since deriving HAND and FIMs at these localized scale does add computational expense.

### 2022-09-15 manuscripts/owp_fim4_2022/sections/discussion.tex
* latest updates * possible resubmission candidate

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Overall, the main observation of this study was how FIM performance can be improved by reducing the Horton-Strahler stream order of the target stream network used for HAND computation.
Most of this change is accounted for by substantially increasing POD and inundation extents in some areas thus reducing FNs.
We believe, as we later argue for, that the increase in POD is primarily driven by an increase in the catchment sizes that is inherently enabled by dividing up the stream network into independent stream networks of unit stream order.
Additionally, we note that reducing drainage order also has some minor influence on reducing inundation extents in other areas and the rate of false alarms.
We believe that this effect is driven by a change in the stage-discharge relationship where a given streamflow now leads to lower river stage values when HAND is computed with target stream networks of unit drainage order.
We seek to explain that these two effects, catchment boundary enlargement and stage-discharge curve lowering, are highly interrelated and cannot be easily detangled.
Lastly, we discuss the diminishing effects on performance that the MS and GMS techniques may have and also any additional effects including enhanced cross-walking abilities.

As evident in the results of the study in Section \ref{sec:results}, a sizable amount of the increase in CSI observed by reducing stream order for HAND computation can be attributed to increases in POD.
This can be inferred by close inspection of Figure \ref{fig:violin_plot} and Table \ref{tab:aggregate_metrics} where changes in POD are significantly higher than that of FAR.
%This intuition is confirmed by previously mentioned work by \citeA{gerapetritis2004behavior} where CSI was shown to be a direct function of both POD and FAR where CSI is maximized when $POD = 1 - FAR$.
Upon investigation of the performance of HAND derived FIM, we note a general increase of catchment sizes for the MS and GMS methods when compared to the FR method as they are now delineated independently of any tributaries that would constrain catchment sizes otherwise.
Additionally, we note significantly less water build up along catchment boundaries especially at the junction of lower order tributaries with lower flow rates and higher order rivers with more flow.
This allows for inundation extents to expand across regions previously encapsulated by catchments of joining reaches in lower flow tributaries.
The water built up along the catchment boundaries can be thought of as a column of water in a cylindrical container (catchment) that has exceeded the elevation of the containers rim.
This, of course, does not represent accurate physics but it is what does occur in HAND when water accumulates along catchment boundaries.

Large scale HUC8 level evaluations can fail to demonstrate fine grain enhancements as they aggregate away many changes that are only clear at more local scales.
Future assessments of OWP FIM should consider finer grain evaluation units as well possible impact assessments using asset information such as building footprints to better illustrate fine grain changes in a more relevant manner to stakeholders.
For now, we offer up Figure \ref{fig:gms_enhancement} which best illustrates the improvement offered by multi-source fluvial flooding capabilities in a more local context.
The figure is comprised of two agreement rasters for two different HAND based FIMs compared to the validation dataset for a given region with a high flow mainstem (500 yr recurrence flow) running horizontally along the region.
Sub-figure \ref{fig:gms_enhancement}a demonstrates the agreement raster for the FR stream network as well as its respective catchment boundary lines symbolized in white and stream network shown in green.
Inspection of this sub-figure denotes a clear spatial pattern where TPs or areas correctly inundated are pooled alongside catchment boundary lines. 
On the other side of the catchment boundary, one can witness large swaths of FNs that should be inundated. 
The FNs also exhibit a spatial pattern as in they tend to collocate within catchments of the pictured mainstems tributaries.
This sort of behavior was introduced early in the paper and shown qualitatively in Figure \ref{fig:catchment_boundaries_issue}.

As an enhancement, this paper proposes computing HAND for stream networks comprised of level-paths independently of one another.
In sub-figure \ref{fig:gms_enhancement}b, the agreement raster for the GMS technique is illustrated as well as the stream network lines in green.
While the entire mosaiced inundation map from GMS (as described in Section \ref{ssec:inundation_mapping} and Equation \ref{eq:hand_fim_depth}) is used to produce this agreement map, we only show the catchments associated with the mainstem of this region that is shown to follow a clear horizontal path.
Showing all the catchments for the tributaries that were all derived independently would convolute the image.
The main message illustrated here is that the catchments associated with the mainstem of this area significantly increase in size.
Since they are not restricted by the catchments of tributaries that lie in the same drainage areas as those of the mainstem, they extend to include those as well.
The consequence for inundation extents is a general increase in spatial coverage of the river's water which shows to be in much better agreement with the benchmark map.
The TPs are no longer bounded by the catchment lines and allowed the expand to their natural extents.

@@

Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged SRC relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's n values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences are very negligible which only leaves room for deviations in volume and bed area.
Again, volume (V(y) or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B(y) or B) is analogous to reach-averaged hydraulic radius but these associations only hold when reach length, L, is considered.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively (see Equation \ref{eq:reach_averaged_mannings_equation}). 
Figures \ref{fig:rating_curve_comparison}b and \ref{fig:rating_curve_comparison}c show how volume and bed area compare across the three methods with GMS having significantly greater values than MS which has greater values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ and plot this ratio against stage, y, to show how these two parameters collectively pull the SRC Q up and changes the SRC accordingly.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.

Much of the increase in inundated pixels, volume, and bed area can be explained by much larger catchments that encompass neighboring tributaries.
These tributaries have a significant amount of bathymetry that is low-lying thus easily included in the geometry for the SRC derivation. 
They also contribute volume and bed area that is technically not perpendicular to the flux of streamflow being accounted for in the stream in question. 
Careful examination of Figure \ref{fig:gms_enhancement}b shows how much larger catchments include neighboring tributaries and the geometry associated with those tributaries. 
This geometry is not perpendicular to the flow that is associated with the main reach thus leading to biases in the SRC.
We consider this fact to have a nuanced effect on skill, while reducing the rate of FPs it also can lead to FNs due to biases in the SRC.

We note that reducing stream order does in turn suffer from diminishing returns in which the increase in mapping skill for applying stream order reduction to roughly 4-5\% of the stream network is about the same as the increase for applying stream order reduction to the remaining 95-96\% of the stream network.
This motivates further work in identifying what the optimal coverage of stream order reduction could be and how to parameterize that coverage. 
One option could be removing stream orders ones and possibly twos and threes from stream order reduction and simply using the inundation from FR from these areas.

Additional careful analysis of Figure \ref{fig:gms_enhancement}a, leads one to note many catchments that don't have inundation or significant inundation.
While the cause of these errors can be varied, we assert here that conflating four networks for use in evaluations leads to significant error.
As one may remember, Section \ref{sssec:cross_walking_networks} details how reach identifiers are conflated for the FIM network back to that of the NWM. 
One of the issues with the FR version of HAND occurs when a reach of given stream order accidentally conflates to that of a neighboring tributary that is of lower order which leads to areas of FNs.
The utilization of MS and GMS only conflates to NWM catchments directly associated with the LP in question which is inherently easy to do with those methods. 
Thus part of the improvement in MS and GMS methods is due to a slight improvement in cross-walking methodology.
The NWM stream network was derived using the NHDPlus V2 dataset which was derived from coarser DEMs than those used here. 
Additional conflation is identified in cross-walking the stream network used by the BLE maps and those of HAND.
Until a singular stream network is used for the NWM, BLE benchmark, and for HAND based FIM, conflation will continue being a source of error.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:discussion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Overall, the main observation of this study was how FIM performance can be improved by reducing the Horton-Strahler stream order of the target stream network used for HAND computation.
Most of this change is accounted for by substantially increasing POD and inundation extents in some areas thus reducing FNs.
We believe, as we later argue, that the increase in POD is primarily driven by an increase in the catchment sizes that is inherently enabled by dividing up the stream network into independent stream networks of unit stream order.
Additionally, we note that reducing drainage order also has some minor influence on reducing inundation extents in other areas and the rate of false alarms.
We believe that this effect is driven by a change in the stage-discharge relationship where a given streamflow leads to lower river stage values when HAND is computed with target stream networks of unit drainage order.
We seek to explain that these two effects, catchment boundary enlargement and stage-discharge curve lowering, are highly interrelated and cannot be easily detangled.
Lastly, we discuss the diminishing effects on performance that the MS and GMS techniques may have and also any additional effects including enhanced cross-walking abilities.

As evident in the results of the study in Section \ref{sec:results}, a sizable amount of the increase in CSI observed by reducing stream order for HAND computation can be attributed to increases in POD.
This can be inferred by close inspection of Figure \ref{fig:violin_plot} and Table \ref{tab:aggregate_metrics} where changes in POD are significantly higher than that of FAR.
%This intuition is confirmed by previously mentioned work by \citeA{gerapetritis2004behavior} where CSI was shown to be a direct function of both POD and FAR where CSI is maximized when $POD = 1 - FAR$.
Upon investigation of the performance of HAND derived FIM, we note a general increase of catchment sizes for the MS and GMS methods when compared to the FR method as they are now delineated independently of any tributaries that would constrain catchment sizes otherwise.
Additionally, we note significantly less water build up along catchment boundaries especially at the junction of lower order tributaries with lower flow rates and higher order rivers with more flow.
This allows for inundation extents to expand across regions previously encapsulated by catchments of joining reaches in lower flow tributaries.
The water built up along the catchment boundaries can be thought of as a column of water in a cylindrical container (catchment) that has exceeded the elevation of the container's rim which does not represent accurate physics.

Large scale HUC8 level evaluations can fail to demonstrate fine grain enhancements as they aggregate away many changes that are only clear at more local scales.
Future assessments of OWP FIM should consider finer grain evaluation units as well possible impact assessments using asset information such as building footprints to better illustrate fine grain changes in a more relevant manner to stakeholders.
For now, we provide Figure \ref{fig:gms_enhancement} which best illustrates the improvement offered by multi-source fluvial flooding capabilities in a more local context.
The figure is comprised of two agreement rasters for two different HAND based FIMs compared to the validation dataset for a given region with a high flow mainstem (500 yr recurrence flow) running horizontally along the region.
Sub-figure \ref{fig:gms_enhancement}a demonstrates the agreement raster for the FR stream network as well as its respective catchment boundary lines symbolized in white and stream network shown in green.
Inspection of this sub-figure denotes a clear spatial pattern where TPs or areas correctly inundated are pooled alongside catchment boundary lines. 
On the other side of the catchment boundary, one can witness large swaths of FNs that should be inundated. 
The FNs also exhibit a spatial pattern as in they tend to collocate within catchments of the pictured mainstems tributaries.
This sort of behavior was introduced early in the paper and shown qualitatively in Figure \ref{fig:catchment_boundaries_issue}.

As an enhancement, this paper proposes computing HAND for stream networks comprised of level-paths independently of one another.
In sub-figure \ref{fig:gms_enhancement}b, the agreement raster for the GMS technique is illustrated as well as the stream network lines in green.
While the entire mosaiced inundation map from GMS (as described in Section \ref{ssec:inundation_mapping} and Equation \ref{eq:hand_fim_depth}) is used to produce this agreement map, we only show the catchments associated with the mainstem of this region that is shown to follow a clear horizontal path.
Showing all the catchments for the tributaries that were all derived independently would convolute the image.
The main message illustrated here is that the catchments associated with the mainstem of this area significantly increase in size.
Since they are not restricted by the catchments of tributaries that lie in the same drainage areas as those of the mainstem, they extend to include those as well.
The consequence for inundation extents is a general increase in spatial coverage of the river's water which shows to be in much better agreement with the benchmark map.
The TPs are no longer bounded by the catchment lines and allowed to expand to their natural extents.

@@

Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged SRC relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's n values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences are very negligible which only leaves room for deviations in volume and bed area.
Again, volume (V(y) or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B(y) or B) is analogous to reach-averaged hydraulic radius but these associations only hold when reach length, L, is considered.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively (see Equation \ref{eq:reach_averaged_mannings_equation}). 
Figures \ref{fig:rating_curve_comparison}b and \ref{fig:rating_curve_comparison}c show how volume and bed area compare across the three methods with GMS having significantly greater values than MS which has greater values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ and plot this ratio against stage, y, to show how these two parameters collectively pull the Q up and changes the SRC accordingly.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.

Much of the increase in inundated pixels, volume, and bed area can be explained by much larger catchments that encompass neighboring tributaries.
These tributaries have a significant amount of bathymetry that is low-lying thus easily included in the geometry for the SRC derivation. 
They also contribute volume and bed area that is technically not perpendicular to the flux of streamflow being accounted for in the stream in question. 
Careful examination of Figure \ref{fig:gms_enhancement}b shows how much larger catchments include neighboring tributaries and the geometry associated with those tributaries. 
This geometry is not perpendicular to the flow that is associated with the main reach thus leading to biases in the SRC.
We consider this to have a nuanced effect on skill, while reducing the rate of FPs it also can lead to FNs due to biases in the SRC.

We note that reducing stream order does suffer from diminishing returns where the increase in mapping skill for applying stream order reduction to roughly 4-5\% of the stream network is about the same as the increase for applying stream order reduction to the remaining 95-96\% of the stream network.
This motivates further work in identifying what the optimal coverage of stream order reduction could be and how to parameterize that coverage. 
One option could be removing lower stream orders (e.g. 1 and 2) from stream order reduction and simply using the inundation from FR from these areas.

Additional analysis of Figure \ref{fig:gms_enhancement}a, reveals that some catchments don't have inundation or significant inundation.
While the cause of these errors can be varied, we assert here that conflating four networks for use in evaluations leads to significant error.
Section \ref{sssec:cross_walking_networks} details how reach identifiers are conflated for the FIM network back to that of the NWM. 
One of the issues with the FR version of HAND occurs when a reach of given stream order accidentally conflates to that of a neighboring tributary that is of lower order which leads to areas of FNs.
The utilization of MS and GMS only conflates to NWM catchments directly associated with the LP in question which is inherently easy to do with those methods. 
Thus part of the improvement in MS and GMS methods is due to a slight improvement in cross-walking methodology.
The NWM stream network was derived using the NHDPlus V2 dataset which was derived from coarser DEMs than those used here. 
Additional conflation is identified in cross-walking the stream network used by the BLE maps and those of HAND.
Until a singular stream network is used for the NWM, BLE benchmark, and for HAND based FIM, conflation will continue being a source of error.

### 2022-09-15 manuscripts/owp_fim4_2022/sections/introduction.tex
* latest updates * possible resubmission candidate

When used for FIM applications, HAND assumes only fluvial inundation sourced from its nearest drainage line is accounted for \cite{nobre2016hand,mcgehee2016modified}.
Catchments are independent of one another for FIM purposes meaning a reaches' stage value is only used to threshold the HAND values within its respective catchment \cite{liu2016cybergis,zheng2018river,zheng2018geoflood}.
This assumption plays to the ``Nearest Drainage'' term in HAND and creates a significant limitation within HAND for FIM applications \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
At the junction of high stream order and high flow rivers with lower flow tributaries, there can be a lack of inundation extents exhibited which is known colloquially in the forecasting community as the ``catchment boundary problem''.
The academic community has somewhat referenced this issue before but it has been characterized more as a problem with the stream delineation process that comes from thresholding the drainage accumulation maps \cite{nobre2016hand,li2020evaluation}.
Later in this study, we will re-introduce this problem and demonstrate how we initialize with a stream network (that of the NWM's) and thus avoid having to threshold accumulations to some arbitrary value to define stream networks.
We illustrate how computing HAND independently for stream lines of unit stream order can significantly enhance FIM performance by accounting for multiple sources of fluvial inundation that may exist in certain regions and flow scenarios.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{HAND Implementations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Due to significant advances in high-performance computing (HPC) and large scale high-resolution DEMs such as the 3D Elevation Program (3DEP) seamless at the 1/3 arc-second (approximately 10 m depending on latitude) scale, HAND has been implemented into software for large-scale, continental computation. 
As part of the OWP's Innovators Program and NWC's Summer Institute, the National Flood Interoperability Experiment (NFIE) generated FIM hydrofabric (will be used interchangeably with the datasets produced by HAND) rapidly on a HPC \cite{maidment2017conceptual,liu2016cybergis}. 
NFIE used open-source dependencies including the Terrain Analysis Using Digital Elevation Models (TauDEM) \cite{tarboton2005terrain} and the Geospatial Data Abstraction Library (GDAL) \cite{warmerdam2008geospatial} to compute HAND for the Continental United States (CONUS) at 331 Hydrologic Unit Code (HUC) 6 processing units in 1.34 CPU years.
By allocating 31 nodes at 20 cores per node for a total of 620 available cores to the overall operation, it enabled the production to finish up in 36 hours consuming 3.2TB of peak memory and 5TB of total disk space.
Originally, NFIE utilized the NHD Medium Resolution (MR) to etch or burn flowlines prior to further conditioning but more recent work has advanced this to the more current NHDPlus High Resolution (NHDPlusHR) which better agrees with the 10 m DEM from the NHDPlusHR program \cite{liu2020height}. 
The original NFIE dataset was employed by the NWC as an unofficial demonstration to produce forecast FIM from the NWM for additional guidance in hydro-blind regions.
Further work by \citeA{djokic2019arc}, implemented a series of improvements to HAND including equidistant reaches, updates to use with NHDPlusHR hydrography, and AGREE DEM reconditioning \cite{hellweger1997agree} into an ESRI Arc-Hydro workflow with use in ArcGIS. 
More notably the software added the ability to derive HAND on both the NWM FR and MS stream networks to consider multiple sources of fluvial inundation along high impact rivers of primary forecasting concern.

Related to these efforts, the United States Geological Survey (USGS) has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses SRCs with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30 m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Office of Water Prediction Flood Inundation Mapping}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In order to mitigate the ever increasing threat of flooding to life and property, an operational capability is required to extend NWM streamflow forecasts to river stages, inundation extents, and inundation depths.
OWP FIM is introduced here as a continental scale capability that generates these products at high spatial and temporal resolutions.
Here we introduce OWP FIM that utilizes a few of the latest techniques in HAND based FIM oriented for use with the NWM in continental scale operational forecasting settings. 
Within the operational framework of OWP FIM, we introduce research demonstrating how FIM performance skill with HAND can be improved by discretizing stream networks into units of an effective unit Horton-Strahler stream order \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} for HAND computation contexts.
Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to the stream accumulation threshold which is closely related to stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
Here we demonstrate how reducing a HAND processing unit's stream network to a singular stream order discretized by level paths (LP), can enhance FIM skill by accounting for multiple possible sources of fluvial inundation.
This capability is introduced progressively as MS (whose network represents about 4\% of FR network) and to a higher degree Generalized Mainstems (GMS) (covers entire FR network) which will be explained later on.
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM with applications.
%

>>>>

When used for FIM applications, HAND assumes only fluvial inundation sourced from its nearest drainage line is accounted for \cite{nobre2016hand,mcgehee2016modified}.
Catchments are independent of one another for FIM purposes meaning a reaches' stage value is only used to threshold the HAND values within its respective catchment \cite{liu2016cybergis,zheng2018river,zheng2018geoflood}.
This assumption plays to the ``Nearest Drainage'' term in HAND and creates a significant limitation within HAND for FIM applications \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation,nobre2016hand}.
At the junction of high stream order and high flow rivers with lower flow tributaries, there can be a lack of inundation extents exhibited which is known colloquially in the forecasting community as the ``catchment boundary problem''.
The academic community has somewhat referenced this issue before but it has been characterized more as a problem with the stream delineation process that comes from thresholding the drainage accumulation maps \cite{nobre2016hand,li2020evaluation}.
Later in this study, we will re-introduce this problem and demonstrate how we initialize with a stream network (that of the NWM's) and thus avoid having to threshold accumulations to some arbitrary value to define stream networks.
We illustrate how computing HAND independently for stream lines of unit stream order can significantly enhance FIM performance by accounting for multiple sources of fluvial inundation that may exist in certain regions and flow scenarios.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{HAND Implementations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Due to significant advances in high performance computing (HPC) and large scale high resolution DEMs such as the 3D Elevation Program (3DEP) seamless at the 1/3 arc-second (approximately 10 m depending on latitude) scale, HAND has been implemented into software for large-scale, continental computation. 
As part of the OWP's Innovators Program and NWC's Summer Institute, the National Flood Interoperability Experiment (NFIE) generated FIM hydrofabric (will be used interchangeably with the datasets produced by HAND) rapidly on a HPC \cite{maidment2017conceptual,liu2016cybergis}. 
NFIE used open-source dependencies including the Terrain Analysis Using Digital Elevation Models (TauDEM) \cite{tarboton2005terrain} and the Geospatial Data Abstraction Library (GDAL) \cite{warmerdam2008geospatial} to compute HAND for the Continental United States (CONUS) at 331 Hydrologic Unit Code (HUC) 6 processing units in 1.34 central processing unit (CPU) years.
By allocating 31 nodes at 20 cores per node for a total of 620 available cores to the overall operation, it enabled the production to finish up in 36 hours consuming 3.2 terrabyte (TB) of peak memory and 5 TB of total disk space.
Originally, NFIE utilized the NHD Medium Resolution (MR) to etch or burn flowlines prior to further conditioning but more recent work has advanced this to the more current NHDPlus High Resolution (NHDPlusHR) which better agrees with the 10 m DEM from the NHDPlusHR program \cite{liu2020height}. 
The original NFIE dataset was employed by the NWC as an unofficial demonstration to produce forecast FIM from the NWM for additional guidance in hydro-blind regions.
Further work by \citeA{djokic2019arc}, implemented a series of improvements to HAND including equidistant reaches, updates to use with NHDPlusHR hydrography, and AGREE DEM reconditioning \cite{hellweger1997agree} into an ESRI Arc-Hydro workflow with use in ArcGIS. 
More notably the software added the ability to derive HAND on both the NWM FR and MS stream networks to consider multiple sources of fluvial inundation along high impact rivers of primary forecasting concern.

Related to these efforts, the United States Geological Survey (USGS) has invested in relative elevation HAND-like methods via work in the GIS Flood Tool (GFT) that also uses SRCs with cross-sections for stage-discharge relationships \cite{verdin2016software}.
Additional investment by \citeA{petrochenkov2020pygft} was able to successfully scale this approach by transitioning the method to open-source Python source code (PyGFT) and implementing novel interpolation methods to help address some of the catchment boundary discontinuities discussed more in this paper.
In addition to the domestic work done in the US, some studies have expanded upon HAND to cover global domains at 30 m resolutions \cite{yamazaki2019merit,donchyts2016global}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Office of Water Prediction Flood Inundation Mapping}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
In order to mitigate the ever increasing threat of flooding to life and property, an operational capability is required to extend NWM streamflow forecasts to river stages, inundation extents, and inundation depths.
OWP FIM is introduced here as a continental scale capability that generates these products at high spatial and temporal resolutions.
Here we introduce OWP FIM that utilizes a few of the latest techniques in HAND based FIM oriented for use with the NWM in continental scale operational forecasting settings. 
Within the operational framework of OWP FIM, we introduce research demonstrating how FIM performance skill with HAND can be improved by discretizing stream networks into units of an effective unit Horton-Strahler stream order \cite{horton1945erosional,strahler1952hypsometric,strahler1952hypsometric} for HAND computation contexts.
Previous authors dating back to the first HAND for FIM work by \citeA{nobre2016hand} have noted a sensitivity of mapping skill to the stream accumulation threshold which is closely related to stream density and the maximum Horton-Strahler stream order (or simply stream order) of the processing unit employed \cite{zhang2018comparative,mcgehee2016modified,li2020evaluation}.
Here we demonstrate how reducing a HAND processing unit's stream network into discrete level paths (LPs) of singular, effective stream order, can enhance FIM skill by accounting for multiple possible sources of fluvial inundation.
This capability is introduced progressively as MS (whose network represents about 4\% of FR network) and to a higher degree Generalized Mainstems (GMS) (covers entire FR network) which will be explained later on.
The following methods and results describe the work in more detail and demonstrate its efficacy in producing enhanced FIM for the NWM.
%
